---
title: "Introduction to mitss"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to mitss}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The `mitss` package implements Multiple Imputation using Two Subclassification Splines (MITSS), a method for estimating causal effects of binary treatments in observational studies. This vignette demonstrates the basic usage and key features of the package.

## Installation

```{r eval=FALSE}
# Install from GitHub
devtools::install_github("jcochanc/mitss")
```

## Load the package

```{r}
library(mitss)
```

## Example 1: Simple Setting

Let's start with a simple example where we have a single covariate that affects both treatment assignment and outcomes.

```{r example1}
# Set seed for reproducibility
set.seed(42)

# Generate data
n <- 1000
X <- rnorm(n)

# Treatment assignment depends on X
e <- plogis(0.5 * X)
W <- rbinom(n, 1, e)

# Outcome depends on treatment and X
# True treatment effect is 1.0
Y <- 1.0 * W + 0.5 * X + rnorm(n, sd = 0.5)

# Estimate treatment effect
result <- mitss(Y, W, X, M = 25, seed = 123)
print(result)
```

The true treatment effect is 1.0, and MITSS recovered it well.

## Example 2: Multiple Covariates

Now let's consider a more realistic setting with multiple covariates:

```{r example2}
set.seed(123)
n <- 800

# Generate three covariates
X <- matrix(rnorm(n * 3), ncol = 3)
colnames(X) <- paste0("X", 1:3)

# Treatment assignment depends on all covariates
e <- plogis(0.3 * X[,1] - 0.4 * X[,2] + 0.2 * X[,3])
W <- rbinom(n, 1, e)

# Outcome with heterogeneous treatment effect
Y <- 2.0 + 
     0.8 * W + 
     0.3 * X[,1] + 
     0.2 * X[,2] - 
     0.1 * X[,3] + 
     rnorm(n)

# Estimate
result2 <- mitss(Y, W, X, M = 25)
summary(result2)
```

## Visualizing Results

The package provides several plotting functions:

```{r plots, fig.width=7, fig.height=5}
# Distribution of imputation estimates
plot(result2, type = "estimates")

# Distribution of propensity scores
plot(result2, type = "propensity")

# Subclass distribution
plot(result2, type = "balance")
```

## Checking for Overlap

It's crucial to check that there is adequate overlap in the propensity score distributions:

```{r overlap}
# Check overlap
overlap_check <- check_overlap(result2$propensity_scores, W)
cat("Jensen-Shannon Divergence:", round(overlap_check$jsd, 3), "\n")
cat("Adequate overlap:", overlap_check$adequate_overlap, "\n")
```

## Trimming for Better Overlap

If overlap is poor, we should trim non-overlapping units:

```{r trimming}
# Example with poor overlap
set.seed(456)
n <- 500
X_poor <- rnorm(n)
e_poor <- plogis(2 * X_poor)  # Strong dependence -> poor overlap
W_poor <- rbinom(n, 1, e_poor)
Y_poor <- W_poor + X_poor + rnorm(n)

# Check overlap
overlap_poor <- check_overlap(e_poor, W_poor)
cat("JSD before trimming:", round(overlap_poor$jsd, 3), "\n")

# Trim
trimmed <- trim_by_propensity(Y_poor, W_poor, matrix(X_poor, ncol=1), e_poor)
cat("Removed", trimmed$n_removed, "units\n")

# Re-estimate on trimmed data
result_trimmed <- mitss(trimmed$Y, trimmed$W, trimmed$X, 
                        e_hat = trimmed$e_hat, M = 25)
print(result_trimmed)
```

## Assessing Covariate Balance

We can check covariate balance:

```{r balance}
# Before matching/subclassification
bias_df <- calculate_standardized_bias(X, W)
print(bias_df)

# Typical threshold: |std_bias| < 0.25 indicates good balance
# MITSS achieves balance through subclassification and orthogonalization
```

## Non-linear Estimands

MITSS can estimate non-linear estimands such as the proportion of units for whom treatment is harmful:

```{r nonlinear}
# Estimate proportion harmed
harmful <- estimate_harmful_proportion(result2)
cat("\nProportion for whom treatment is not beneficial:\n")
cat("  Estimate:", round(harmful$prop_harmful, 3), "\n")
cat("  95% CI: [", round(harmful$ci_lower, 3), ",", 
    round(harmful$ci_upper, 3), "]\n")
```

## Comparison with Naive Estimator

Let's compare MITSS with a naive difference-in-means estimator:

```{r comparison}
# Naive estimator (ignores confounding)
naive_est <- mean(Y[W==1]) - mean(Y[W==0])

# MITSS estimator
mitss_est <- result2$estimate

cat("Naive estimate:     ", round(naive_est, 3), "\n")
cat("MITSS estimate:     ", round(mitss_est, 3), "\n")
cat("True effect:        ", 0.8, "\n")
cat("\nMITSS bias:         ", round(mitss_est - 0.8, 3), "\n")
cat("Naive bias:         ", round(naive_est - 0.8, 3), "\n")
```

## Sensitivity to Number of Imputations

The number of imputations M affects precision:

```{r sensitivity}
# Compare different M values
M_values <- c(5, 10, 25, 50)
results_by_M <- lapply(M_values, function(m) {
  r <- mitss(Y, W, X, M = m, seed = 999)
  c(estimate = r$estimate, se = r$std_error, width = r$ci_upper - r$ci_lower)
})

results_df <- do.call(rbind, results_by_M)
rownames(results_df) <- paste0("M=", M_values)
print(round(results_df, 4))

# Generally, M=25 is sufficient for most applications
```

## Summary

This vignette demonstrated:

1. Basic usage of `mitss()` for estimating treatment effects
2. Checking and improving propensity score overlap
3. Assessing covariate balance
4. Estimating non-linear estimands
5. Visualizing results
6. Comparing with naive estimators

For more details on the theoretical foundation, see Gutman and Rubin (2015).

## References

Gutman, R. and Rubin, D.B. (2015). Estimation of causal effects of binary treatments in unconfounded studies. *Statistics in Medicine*, 34(26), 3381-3398. doi:10.1002/sim.6532
